\chapter{Rhythmic recitation exercises}

The following exercises are meant to help you master solkattu recitation. Check-points
are in bold face, bar where one beat ends, simple brackets denote one complete sol.

\begin{enumerate}
  \item Start by counting tha ka dhi mi per clap (walk is 4 atoms per beat). Don't
    alter the walk. But now, start saying tha ka tha ki ta.
    \begin{center}\ttfamily
    (Tha ka tha ki | ta) (tha ka tha | ki ta) (tha ka | tha ki ta) (tha | ka tha ki ta)
    \end{center}
    Math: there are 4 tha-ka-tha-ki-tas in here. 4 x 5 = 20 atoms. That is 5 claps.

  \item Start by counting tha ka tha ki ta per clap (walk is 5 atoms per beat). Without
    altering the walk start saying, tha ka dhi mi.
    \begin{center}\ttfamily
    (Tha ka dhi mi) (tha | ka dhi mi) (tha ka | dhi mi) (tha ka dhi | mi) (tha ka dhi mi)
    \end{center}
    Math: there are 5 tha-ka-dhi-mi\,s in here. 5 x 4 = 20 atoms. That is 4 claps. Note
    that the duration of one clap is unchanged in the above exercise, this one and all
    that follow.

  \item Start by counting tha ki ta per clap (walk is 3 atoms per beat). Hold the speed
    and start saying, tha ka dhi mi.
    \begin{center}\ttfamily
    (Tha ka dhi | mi) (tha ka | dhi mi) (tha | ka dhi mi) (4 claps).
    \end{center}

  \item Start by counting tha ki ta per clap (walk is 3 atoms per beat). Hold the speed
    and start saying, tha ka tha ki ta.
    \begin{center}\ttfamily
    (Tha ka tha | ki ta) (tha | ka tha ki | ta) (tha ka | tha ki ta)
    \end{center}

  \item Start by counting tha ki ta tha ki ta per clap (walk is 6 atoms per beat). Hold
    the speed and start saying, tha ka dhi mi tha ki ta.
    \begin{center}\ttfamily
    (Tha ka dhi mi tha ki | ta) (tha ka dhi mi tha | ki ta) (tha ka dhi mi | tha ki ta)
    (tha ka dhi | mi tha ki ta) (Tha ka | dhi mi tha ki ta) (tha | ka dhi mi tha ki ta)
    (7 claps)
    \end{center}
\end{enumerate}

\section*{Practising conclusive fillers in multiple walks}
\addcontentsline{toc}{section}{Practising conclusive fillers in multiple walks}

It is highly useful to be able to recite the conclusive fillers in Chatusram and
Tisram. Practising this is quite easy, when you discover the trick. The trick is
presented below.

\begin{enumerate}
  \item Tha dhi ki ta thom is 5 atoms long. Two of these is 10 atoms long.
    Interestingly, one avartana (one cycle) of Kanda Chapu (2.5-clap cycle) is also 10
    atoms long if the walk is 4 atoms per clap.
    \[ 2.5 \text{ claps} \times 4 \text{ atoms/clap} = 10 \text{ atoms} \]
    Hence, one could recite 2 tha dhi ki ta thoms over one cycle of Kanda Chapu at 4
    atoms per clap --- chatusram. Also,
    \[ 2.5 \text{ claps} \times 6 \text{ atoms/clap} = 15 \text{ atoms} \]
    So, one could recite 3 tha dhi ki ta thom over one cycle of Kanda Chapu at 6 atoms
    per clap --- tisram double speed. Always start with the basic groove for a walk,
    then switch.

  \item Tha dhee - ki ta thom is 6 atoms long. Two of these is 12 atoms long.
    Interestingly, one avartana (one cycle) of Roopaka Chapu (3-clap cycle) is also 12
    atoms long if the walk is 4 atoms per clap.
    \[ 3 \text{ claps} \times 4 \text{ atoms/clap} = 12 \text{ atoms} \]
    Hence, one could recite 2 tha dhee - ki ta thoms over one cycle of Roopaka Chapu at
    4 atoms per clap --- chatusram. Also,
    \[ 3 \text{ claps} \times 6 \text{ atoms/clap} = 18 \text{ atoms} \]
    So, one could recite 3 tha dhee - ki ta thoms over one cycle of Roopaka Chapu at 6
    atoms per clap --- tisram double speed.

  \item Tha - dhee - ki ta thom is 7 atoms long. Two of these is 14 atoms long.
    Interestingly, one avartana (one cycle) of Misra Chapu (3.5-clap cycle) is also 14
    atoms long if the walk is 4 atoms per clap.
    \[ 3.5 \text{ claps} \times 4 \text{ atoms/clap} = 14 \text{ atoms} \]
    Hence, one could recite 2 Tha - dhee - ki ta thoms over one cycle of Misra Chapu at
    4 atoms per clap --- chatusram. Also,
    \[ 3.5 \text{ claps} \times 6 \text{ atoms/clap} = 21 \text{ atoms} \]
    So, one could recite 3 Tha - dhee - ki ta thoms over one cycle of Misra Chapu at 6
    atoms per clap --- tisram double speed.

  \item Follow the above steps for the 8-atom, 9-atom and 10-atom tha dhee ki ta thoms
    and be eternally happy.
\end{enumerate}
